---
title: "Run Canek on a toy example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Canek on a toy example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Canek)

# Functions
## Function to plot the pca coordinates
plotPCA <- function(pcaData = NULL, label = NULL, legPosition = "topleft"){
  col <- as.integer(label) 
  plot(x = pcaData[,"PC1"], y = pcaData[,"PC2"],
       col = as.integer(label), cex = 0.75, pch = 19,
       xlab = "PC1", ylab = "PC2")
  legend(legPosition,  pch = 19,
         legend = levels(label), 
         col =  unique(as.integer(label)))
}
```

## Load the data

On this toy example we use the two simulated batches included in the `SimBatches` data from Canek's package. `SimBatches` is a list containing:

* `batches`: Simulated scRNA-seq datasets with genes (rows) and cells (columns). Simulations were performed using [Splatter](https://bioconductor.org/packages/devel/bioc/vignettes/splatter/inst/doc/splatter.html).
* `cell_type`: a factor containing the celltype labels of the batches

```{r}
lsData <- list(B1 = SimBatches$batches[[1]], B2 = SimBatches$batches[[2]])
batch <- factor(c(rep("Batch-1", ncol(lsData[[1]])),
                  rep("Batch-2", ncol(lsData[[2]]))))
celltype <- SimBatches$cell_types
table(batch)
table(celltype)
```

## PCA before correction

We perform the Principal Component Analysys (PCA) of the joined datasets and scatter plot the first two PCs. The batch-effect causes cells to group by batch. 

```{r}
data <- Reduce(cbind, lsData)
pcaData <- prcomp(t(data), center = TRUE, scale. = TRUE)$x
```

```{r}
plotPCA(pcaData = pcaData, label = batch, legPosition = "bottomleft")
plotPCA(pcaData = pcaData, label = celltype, legPosition = "bottomleft")
```

## Run Canek

We correct the toy batches using the function *RunCanek*. This function accepts:

* List of matrices
* Seurat object
* List of Seurat objects
* SingleCellExperiment object
* List of SingleCellExperiment objects

On this example we use the list of matrices created before.

```{r}
data <- RunCanek(lsData)
```

## PCA after correction

We perform PCA of the corrected datasets and plot the first two PCs. After correction, the cells group by their corresponding cell type.

```{r}
pcaData <- prcomp(t(data), center = TRUE, scale. = TRUE)$x
```

```{r}
plotPCA(pcaData = pcaData, label = batch, legPosition = "topleft")
plotPCA(pcaData = pcaData, label = celltype, legPosition = "topleft")
```

## Session info

```{r}
sessionInfo()
```



