---
title: "Correct SCTransform-normalized Seurat data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correct SCTransform-normalized Seurat data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

`SCTransform` is normally fit *separately per batch* before merging — that's
Seurat's own recommended way to run it ahead of integration. That means each
batch's residuals live on a different, non-comparable scale. `RunCanek()`'s
default `correctEmbeddings = TRUE` corrects in PCA-embedding space, and by
default reuses whatever `"pca"` reduction already exists on the object
instead of computing its own (see
[Correct log-normalized Seurat data](seurat.html) for that default path). For
SCTransform-normalized data specifically, we should reconciled the per-batch
normalizations so that the batch-to-batch scale mismatch is accounted for in the
correction. This vignette shows the reconciliation Seurat itself provides
for exactly this problem, and how `RunCanek()` uses it.

```{r setup}
library(Canek)
library(Seurat)
```

# Create and SCTransform-normalize per batch

We use the two simulated batches included in `SimBatches`, normalizing each
one independently with `SCTransform` before merging — mirroring how you'd
normally run it ahead of integration.

```{r}
x <- lapply(names(SimBatches$batches), function(batch) {
  CreateSeuratObject(SimBatches$batches[[batch]], project = batch)
})
names(x) <- names(SimBatches$batches)
x <- lapply(x, SCTransform, verbose = FALSE)
```

# What happens without reconciling first

If we just merge the independently-normalized batches and call `RunCanek()`
without computing a joint PCA on reconciled data first, it stops with an
error rather than silently computing a PCA on data that isn't actually
comparable across batches:

```{r, error = TRUE}
naive <- merge(x[[1]], x[[2]])
DefaultAssay(naive) <- "SCT"
RunCanek(naive, "orig.ident")
```

# Reconciling the per-batch models

`RunCanek()`'s error points at the fix: select integration features across
the independent models, reconcile the batches' residuals onto a comparable
scale for those features (`PrepSCTIntegration`), merge, and compute a joint
PCA on the reconciled data.

```{r}
features <- SelectIntegrationFeatures(x, nfeatures = 3000)
x <- PrepSCTIntegration(x, anchor.features = features)

x <- merge(x[[1]], x[[2]])
x[["cell_type"]] <- SimBatches$cell_types
DefaultAssay(x) <- "SCT"
VariableFeatures(x) <- features
x <- RunPCA(x, verbose = FALSE)
```

## UMAP before correction

```{r}
x <- RunUMAP(x, reduction = "pca", dims = 1:ncol(Embeddings(x, "pca")),
             reduction.name = "umap_uncorrected", verbose = FALSE)
DimPlot(x, reduction = "umap_uncorrected", group.by = "orig.ident")
DimPlot(x, reduction = "umap_uncorrected", group.by = "cell_type")
```

# Run Canek

With a `"pca"` reduction now present on properly reconciled data,
`RunCanek()` reuses it directly instead of recomputing one — the SCT assay
is auto-detected, so we don't need to pass `assay = "SCT"` explicitly.

```{r}
x <- RunCanek(x, "orig.ident")
Reductions(x)
ncol(Embeddings(x, "canek"))
```

## UMAP after correction

As with the log-normalized workflow, downstream steps should use the
`"canek"` reduction directly rather than re-running `ScaleData`/`RunPCA` on
the SCT assay.

```{r}
x <- RunUMAP(x, reduction = "canek", dims = 1:ncol(Embeddings(x, "canek")),
             reduction.name = "umap_corrected", verbose = FALSE)
DimPlot(x, reduction = "umap_corrected", group.by = "orig.ident")
DimPlot(x, reduction = "umap_corrected", group.by = "cell_type")
```

## Session info

```{r}
sessionInfo()
```
