---
title: "Correct log-normalized Seurat data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correct log-normalized Seurat data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows Canek's default workflow on standard, log-normalized Seurat
data. If your data is SCTransform-normalized instead, see
[Correct SCTransform-normalized Seurat data](SCTransform.html) — SCTransform
fit per batch produces residuals on different scales, so it needs an extra
reconciliation step before Canek can correct it. For the older,
gene-expression-space correction path (`correctEmbeddings = FALSE`), see
[the legacy version of this vignette](seurat_legacy.html).

```{r setup}
library(Canek)
library(Seurat)
```

# Create a Seurat object

We use the two simulated batches included in `SimBatches`.

```{r}
x <- lapply(names(SimBatches$batches), function(batch) {
  CreateSeuratObject(SimBatches$batches[[batch]], project = batch)
})
x <- merge(x[[1]], x[[2]])
x[["cell_type"]] <- SimBatches$cell_types
x
```

```{r}
table(x$orig.ident)
```

# Standard preprocessing

We follow Seurat's standard log-normalization workflow on the whole merged
object.

```{r}
x <- NormalizeData(x, verbose = FALSE)
x <- FindVariableFeatures(x, nfeatures = 100, verbose = FALSE)
x <- ScaleData(x, verbose = FALSE)
x <- RunPCA(x, verbose = FALSE)
```

## UMAP before correction

```{r}
x <- RunUMAP(x, reduction = "pca", dims = 1:ncol(Embeddings(x, "pca")),
             reduction.name = "umap_uncorrected", verbose = FALSE)
DimPlot(x, reduction = "umap_uncorrected", group.by = "orig.ident")
DimPlot(x, reduction = "umap_uncorrected", group.by = "cell_type")
```

Cells separate by `orig.ident` in a way that doesn't line up with `cell_type`
— that's the batch effect we want to correct.

# Run Canek

We pass the column containing the batch information. By default,
`RunCanek()` corrects in PCA-embedding space (`correctEmbeddings = TRUE`)
rather than on gene expression directly. Since `x` already has a `"pca"`
reduction from the step above, Canek infers the number of dimensions to use
from it and reuses that embedding directly instead of recomputing its own
PCA.

```{r}
x <- RunCanek(x, "orig.ident")
Reductions(x)
ncol(Embeddings(x, "canek"))
```

The correction is stored as a new `"canek"` reduction, not a new assay — the
original `RNA` assay is untouched.

```{r}
Assays(x)
```

By default, `RunCanek()` also repeats the correction up to 5 times
(`maxLoop = 5`), feeding each pass's result into the next and stopping early
once the correction stops changing much (`loopTol`). Pass `maxLoop = 1` for
a single pass, or `debug = TRUE` to inspect the per-pass correction
magnitude (`x@tools$RunCanek`).

## UMAP after correction

Downstream steps should use the `"canek"` reduction directly, not re-run
`ScaleData`/`RunPCA` on the original assay — the correction lives in the
reduction, not in the assay's expression values.

```{r}
x <- RunUMAP(x, reduction = "canek", dims = 1:ncol(Embeddings(x, "canek")),
             reduction.name = "umap_corrected", verbose = FALSE)
DimPlot(x, reduction = "umap_corrected", group.by = "orig.ident")
DimPlot(x, reduction = "umap_corrected", group.by = "cell_type")
```

## Session info

```{r}
sessionInfo()
```
