---
title: "Run Canek on Seurat objects (gene-expression-space correction)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Canek on Seurat objects (gene-expression-space correction)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> This is the original content of the Seurat vignette, from before
> `correctEmbeddings` existed. `RunCanek()` now corrects in PCA-embedding
> space by default, returning a `"canek"` reduction rather than a `"Canek"`
> assay — see [Log-normalized data](seurat.html) or
> [SCTransform-normalized data](SCTransform.html) for the current, default
> workflow. The gene-expression-space path shown below still works and is
> still supported, just no longer the default, so the one `RunCanek()`
> call here has been pinned to `correctEmbeddings = FALSE` to keep this page
> runnable as originally written.

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Canek)
library(Seurat)
```

# Create Seurat object

```{r}
x <- lapply(names(SimBatches$batches), function(batch) {
  CreateSeuratObject(SimBatches$batches[[batch]], project = batch) 
})
x <- merge(x[[1]], x[[2]])
x[["cell_type"]] <- SimBatches$cell_types
x
```

```{r}
table(x$orig.ident)
```

```{r}
x <- NormalizeData(x)
x <- FindVariableFeatures(x, nfeatures=100)
VariableFeaturePlot(x)
```

```{r}
x <- ScaleData(x)
```

```{r}
x <- RunPCA(x)
```

## PCA plot before correction

```{r}
DimPlot(x, group.by = "orig.ident")
DimPlot(x, group.by = "cell_type")
```

# Run Canek

We pass the column containing the batch information.

```{r}
x <- RunCanek(x, "orig.ident", correctEmbeddings = FALSE)
```

The features selected during integration are assigned as variable features.

```{r}
head(VariableFeatures(x, assay="Canek"))
```

We scale features and run PCA.

```{r}
x <- ScaleData(x)
x <- RunPCA(x)
```

## PCA plot after correction

```{r}
DimPlot(x, group.by = "orig.ident")
DimPlot(x, group.by = "cell_type")
```
