---
title: "Run Canek on SingleCellExperiment objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Canek with Bioconductor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we demonstrate how to use `Canek` to correct batch effects from `SingleCellExperiment` objects following the recommendations from [Bioconductor](https://bioconductor.org/books/release/OSCA/) multi-sample analysis. 
The toy data sets used in this vignette are included in the `Canek` R package. 

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

## Init

```{r setup, message=FALSE}
library(Canek)
library(SingleCellExperiment)
library(scater)
library(batchelor)
library(scran)
library(ggplot2)
library(patchwork)
```

## Create SingleCellExperiment objects

```{r}
#Create independent SingleCellExperiment objects
sce <- lapply(names(Canek::SimBatches$batches), function(batch) {
  counts <- SimBatches$batches[[batch]]     #get the counts
  SingleCellExperiment(list(counts=counts),      #create the sce object
                       mainExpName = "example_sce")
})

#Add batch labels
names_batch <- c("B1", "B2")
names(sce) <- names_batch
sce$B1[["batch"]] <- "B1"
sce$B2[["batch"]] <- "B2"

#Add cell-type labels (included in Canek's package)
celltypes <- Canek::SimBatches$cell_types
ct_B1_idx <- 1:ncol(sce$B1)
ct_B2_idx <- (ncol(sce$B1)+1):length(celltypes)
sce$B1[["celltype"]] <- celltypes[ct_B1_idx]
sce$B2[["celltype"]] <- celltypes[ct_B2_idx]
```

Let's check the cells distribution across batches.

```{r}
table(sce$B1$batch)
table(sce$B2$batch)
```

## Pre-processing

In the pre-processing steps, we perform the normalization of counts and the variance analysis of genes independently for each batch. Then, we find common variable genes to use in downstream analyses. 

### Normalization

We perform scaling normalization within each batch using the `multiBatchNorm` function from [batchelor](https://bioconductor.org/packages/release/bioc/html/batchelor.html) Bioconductor package. 

```{r}
sce <- batchelor::multiBatchNorm(sce[["B1"]], sce[["B2"]])
names(sce) <- names_batch
```

### Feature selection

We use the `combineVar` function from [scran](https://bioconductor.org/packages/release/bioc/html/scran.html) Bioconductor package. This function receives independent variance models of genes for each of the batches. This allows us to select highly variable genes while preserving within-batch differences. 

We found 273 combined variable features.

```{r}
#Get independent features' variant models
sce_var <- lapply(X = sce, FUN = scran::modelGeneVar) 
#Combine the independent variables
combined_features <- scran::combineVar(sce_var[["B1"]], sce_var[["B2"]])
#Select features using the `bio` statistics (see combineVar documentation for further details)
combined_features <- combined_features$bio > 0
sum(combined_features)
```

### Visualization before batch effects correction

Then, we merge the two datasets and use the combined variable features for the Principal Components Analysis.

```{r}
uncorrected <- cbind(sce[["B1"]], sce[["B2"]])
uncorrected <- scater::runPCA(x = uncorrected[combined_features,], scale = TRUE)
```

Let's check out the elbow plot of the PCs.

```{r}
pca_uncorrected <- SingleCellExperiment::reducedDim(x = uncorrected, type = "PCA")
variance <- apply(X = pca_uncorrected, MARGIN = 2, FUN = var)
plot(x = 1:30, y = variance[1:30], xlab = "Principal components", ylab = "Variance")
```

The variance stabilizes after the first seven PCs. In this test, we will use 10 PCs to calculate the UMAP visualization, but feel free to try other numbers.

```{r}
set.seed(777)
uncorrected <- scater::runUMAP(x = uncorrected, dimred = "PCA", pca = 10)
```

We can now visualize the uncorrected data by batch and cell-type labels. 

```{r}
p1 <- scater::plotUMAP(object = uncorrected, colour_by="batch") 
p2 <- scater::plotUMAP(object = uncorrected, colour_by="celltype") 
p1 + p2
```

We can observe that the cells labeled as *Cell Type 1* got divided into two groups that correlated with batch labels. Let's minimize this batch differences with Canek.

## Correcting batch effects with Canek

Canek accepts a `SingleCellExperiments` object with a `batch` identifier. We can pass a vector of variable features to use in the integration.

```{r}
features <- rownames(uncorrected)[combined_features]
#RunCanek in a list of SingleCellExperiment objects
#corrected <- Canek::RunCanek(x = sce, features = features)
#RunCanek in a single object with a batch identifier
corrected <- Canek::RunCanek(x = uncorrected, batches = "batch", features = features)
```

### Visualization after batch effects correction with Canek

To perform PCA it's important to use the corrected log counts. These are saved in the assay `Canek` in the `SingleCellExperiment` object and can be specified by changing the `exprs_values` parameter.

```{r}
corrected <- scater::runPCA(x = corrected, scale = TRUE, exprs_values = "Canek")
```

Let's check out the elbow plot after correction. 

```{r}
pca_corrected <- SingleCellExperiment::reducedDim(x = corrected, type = "PCA")
variance <- apply(X = pca_corrected, MARGIN = 2, FUN = var)
plot(x = 1:30, y = variance[1:30], xlab = "Principal components", ylab = "Variance")
``` 

After correction, the elbow plot has changed and 5 PCs will be enough to calculate the UMAP visualization.

```{r}
set.seed(777)
corrected <- scater::runUMAP(x = corrected, dimred = "PCA", pca = 5)
```

We can now visualize the corrected data by batch and cell-type labels. We can observe that after batch-correction with Canek, we minimize the batch difference in cells from *Cell Type 1* while preserving the other cell types.

```{r}
p1 <- scater::plotUMAP(object = corrected, colour_by="batch")
p2 <- scater::plotUMAP(object = corrected, colour_by="celltype")
p1 + p2
```
## Session info

```{r}
sessionInfo()
```

